The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-12 article.tex
Ready for @rsetienne to review the methods

% Methods

@@

From each biological parameter set, a protracted birth-death tree is simulated,
with the same crown age as [Etienne et al., 2014] of 15 million years. 
Each protracted birth-death tree uses a different random number
generatior seed, and thus will be unique, resulting in a balanced 
data set.

From an incipient species tree, we sample a species tree
by picking either the youngest or oldest
species-to-be to represent the full species.
We use either sampling method in 50\% of all cases. We chose not to use
a random species-to-be to represent a full species as to maximize variance
in this aspect. Note that only in few cases (when speciation completion rate
is low) the sampling method has an effect on the branch length distribution
of the sampled species tree, as this requires a very specific scenario.

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we can use this exact site model as a prior, which we 
know is the truth.
The DNA sequence length and mutation rate was chosen as such to provide a
resolution of a 1000 years, that is, to have one expected nucleotide change 
per 1000 years per lineage on average. As our time unit is in 
millions of years, and our crown age is 15 million years,
this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K (which is
coincidentally the same as \cite{moller2018}). As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior later.

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model with 
the same fixed rate as used in the simulation of the alignment (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC sampling interval is 
chosen as such that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC sampling interval is doubled up until
the ESS exceeds 200.

@@

Each posterior's phylogeny is compared to the (sampled) species tree.
To quantify the difference between two phylogenies, the nLTT statistic
is used. The nLTT statistic equates to the area between the normalized
lineages-through-time-plots of two phylogenies, which will be a value between 
zero (for identical phylogenies) and one. As the nLTT statistic increases
when phylogenies get increasingly different, we use the nLTT statistic
as a measure of inference error. Comparing the one (sampled) species tree
with each of the posterior's species trees, a distribution nLTT statistics
is created.

We predict that nLTT statistic values increase 
with an increasing protractednedness (that is, a low speciation completion rate),
but we cannot predict the the extent of this error, as it has never been 
measured [NOTE: this may be easy to change: Etienne \& Rosindel 2012 show the equations
for the expected number of lineages through time. I think I may be able to pull
of a prediction here, by fitting a BD curve on a PBD curve with a minimized error,
but research would be needed to investigate that].

There will be two data sets produced by this research.
The first data set will be a general balanced data set to chart
the effect of the biological parameters on the nLTT statistic
distribution. The second data set charts the effect of sampling
subspecies (young versus old) and only uses phylogenies in which this sampling
has an effect.

Each data set will consist out of two files, containing raw and summarized 
data respectively. We placed 
an upper limit for this data set's size at the maximum size of a table (called 
a data frame) of the R programming language, in which such a table may 
have maximally $2^{32}-1$ cells and occupy less then 2 gigabytes in memory.
The raw data file will contain parameters and 1000 nLTT statistics values.
Each row will contain the parameter set and these nLTT statistics.
From the constraints of the R programming language, this allows for 40K rows. 
The file with the summarized data will provide summary statistics of the raw
data. Each row will contain a parameter set, summary statistics and other
measurements. See the appendix, chapter 'Data set specifications', for the full list.

Our results will show the general effect of the biological parameters (speciation initiation, 
speciation completion and extinction rate) using the balanced data set,
and the effect of sampling using the data set conditioned on sampling having an effect.
In both cases, the nLTT statistics distribution will be plotted per
biological parameter set using a violin plot. A violin plot is chosen to
maintain information about the distribution. When displaying the effect of a 
sampling regime, per regime, the nLTT statistics distribution is shown per
biological parameter set.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set.
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters \\
    $1000$ & nLTT statistic \\
    \hline
    $12$ & simulation parameters \\
    $1$ & expected duration of speciation (from Etienne \& Rosindell 2012) \\
    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
    $1$ & number of taxa in species tree (also: number of species) \\
    $1$ & mean nLTT statistic \\
    $1$ & median nLTT statistic \\
    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. $n$ denotes the number
    of columns a certain item will occupy. Above the horizontal line,
    the content of the raw data file is specified, resulting in a table of 
    1012 columns and 40K rows. Below the horizontal line,
    the content of the summary data file specified, resulting in a table of
    40 columns and 40K rows.
  }
  \label{table:raw_data_set_specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods (but we are not allowed to keep this header)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

From each biological parameter set, a protracted birth-death tree is simulated,
using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
with the same crown age as \cite{etienne2014} of 15 million years. 
Each protracted birth-death tree uses a different random number
generatior seed, and thus will be unique, resulting in a balanced 
data set.

From an incipient species tree, we sample a species tree. To do
so, from each species a sub-species must be chosen to represent
the good species as a whole. To clarify, it may be that an
incipient species branched of from a good species. Both of these
subspecies are recognized as the good species, and both can be 
picked as a representative. Here we use three sampling scenario's,
in which we pick the youngest (most recent incipient speciation), 
oldest or random subspecies. Note that only in few 
cases (when speciation completion rate is low) the incipient species
chosen will have an effect on the branch length distribution of species 
tree, as this requires a very specific scenario.

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny, using the phangorn package (\cite{phangorn}). 
The nucleotides of the DNA alignment follow a Jukes-Cantor (\cite{jc69})
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we use this exact site model as the (obviously correct) site model prior.
The DNA sequence length and mutation rate are chosen as such to provide a
resolution of a 1000 years, that is, to have one expected nucleotide change 
per 1000 years per lineage on average. As we use a crown age is 15 million years,
this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
changes per million years per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as the expected number of mutations (including those
that do not modify the nucleotide), which equals the
nucleotide substitution rate times the crown age, which equals 15000 nucleotides (which is
coincidentally the same as \cite{moller2018}). As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior in the Bayesian inference.

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2 \cite{beast2} using the 
pirouette (\cite{pirouette}) package. As our site model, we assume a Jukes-Cantor 
nucleotide substitution model, as we used that in the simulation of the alignment.
For our clock model, we assume a strict clock with the same fixed rate as 
used in the simulation of the alignment. The tree prior assumed is the BD model, 
as this simplification is the goal of this research. 
Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an effective sample size (ESS) of the posterior of at least 200 
to have a strong enough inference (\cite{beastbook}). The MCMC sampling interval is 
chosen in such a way that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC sampling interval is doubled up until
the ESS exceeds 200.

@@

Each posterior's phylogeny is compared to the (sampled) species tree
by the nLTT statistic (\cite{janzen2015}), using the nLTT package (\cite{nltt}). 
The nLTT statistic equates to the area between the normalized
lineages-through-time-plots of two phylogenies, which will be a value between 
zero (for identical phylogenies) and one. As the nLTT statistic increases
when phylogenies get increasingly different, we use the nLTT statistic
as a measure of inference error. Comparing the one (sampled) species tree
with each of the posterior's species trees, a distribution nLTT statistics
is created.

There will be two data sets produced by this research.
The first data set will be a general balanced data set to chart
the effect of the biological parameters on the nLTT statistic
distribution. In this data set, incipient species are sampled 
randomly to represent a good species.
The second data set charts the effect of sampling
subspecies and only uses PBD trees in which this sampling
has an effect. For each of these trees, we sample both youngest and
oldest subspecies. We predict that these two most extreme sampling methods
result in the most pronounced differences.

Each data set will be stored as a comma-seperated file. We placed 
an upper limit for this data set's size at the maximum size of a table (called 
a data frame) of the R programming language, in which such a table may 
have maximally $2^{32}-1$ cells and occupy less then 2 gigabytes in memory.
Each row will contain the parameter set, nLTT statistics and some 
diagnostics [NOTE: should I keep the diagnostics?] (see \ref{table:specs} for the 
complete list). From the constraints of the R programming language, this allows for 40K rows.

Our results will show the general effect of the biological parameters (speciation initiation, 
speciation completion and extinction rate) using the balanced data set,
and the effect of sampling using the data set conditioned on sampling having an effect.
In both cases, the nLTT statistics distribution will be plotted per
biological parameter set using a violin plot. A violin plot is chosen to
maintain information about the distribution. When displaying the effect of a 
sampling regime, per regime, the nLTT statistics distribution is shown per
biological parameter set.
We predict that nLTT statistic values increase 
with an increasing protractednedness (that is, a low speciation completion rate),
but we cannot predict the the extent of this error, as it has never been 
measured.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Random \\
    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set.
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters, see table \ref{table:parameters} \\
    $1000$ & nLTT statistic values \\
    $1$ & [NOTE: keep this in?] expected duration of speciation (from Etienne \& Rosindell 2012) \\
    $1$ & [NOTE: keep this in?] number of taxa in incipient species tree (also: number of subspecies) \\
    $1$ & [NOTE: keep this in?] number of taxa in species tree (also: number of species) \\
    $1$ & [NOTE: keep this in?] mean nLTT statistic \\
    $1$ & [NOTE: keep this in?] median nLTT statistic \\
    $1$ & [NOTE: keep this in?] lower bound of 95\% HPD value of nLTT statistic \\
    $1$ & [NOTE: keep this in?] upper bound of 95\% HPD value of nLTT statistic \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. $n$ denotes the number
    of columns a certain item will occupy, resulting in a table of 
    1030 columns and 40K rows.
  }
  \label{table:specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
